# Data and data management II

```{r}
#| label: setup-03
#| include: false
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
options(width = 78)
```

Last week we did data management with base R. This week we do the same work with the **tidyverse**, a collection of packages built around a common idea: a small number of verbs, each doing one thing, joined together into a sequence.

By the end of this chapter you will have built the analysis data set that the rest of the course uses.

```{r}
library(dplyr)
raw <- read.csv("data/cses6_swe_raw.csv")
```

`dplyr` is the part of the tidyverse that handles data frames. You can load the whole collection with `library(tidyverse)` instead, which also brings in `ggplot2` for session 5 and a few others.

> When you load `dplyr` you will see a message about objects being **masked**. `dplyr` has functions called `filter()` and `lag()`, and so does base R. The message tells you which version wins. This is normal and not a problem — but it is the reason a function can behave differently depending on what you have loaded.

## The pipe

The one piece of syntax to learn first. Consider a small nested calculation:

```{r}
round(mean(c(4, 8, 15, 16), na.rm = TRUE), digits = 1)
```

To read that you have to start in the middle, at `c()`, work outwards to `mean()`, then outwards again to `round()`. The order you read is the reverse of the order things happen.

The **pipe**, written `|>`, takes what is on its left and passes it as the first argument to what is on its right.

```{r}
c(4, 8, 15, 16) |>
  mean(na.rm = TRUE) |>
  round(digits = 1)
```

Same answer, but now the reading order and the doing order are the same: take these numbers, take their mean, round it. Read `|>` as "and then".

The keyboard shortcut in RStudio is **Ctrl+Shift+M** (**Cmd+Shift+M** on a Mac).

> You will also see `%>%`, an older pipe from the `magrittr` package that the tidyverse used before R had one of its own. They do the same thing here. `|>` is built into R and needs no package, so it is the one to use.

This matters because of principle 2.1 of how these materials are written: operations should not be nested inside one another. The pipe is how you avoid nesting without inventing a name for every intermediate step.

## The verbs

Almost all data management is five operations. Each has a verb, each takes a data frame as its first argument, and each gives a data frame back — which is what makes them combine.

| Verb | Does |
|---|---|
| `select()` | choose columns |
| `filter()` | choose rows |
| `mutate()` | make new columns |
| `rename()` | change column names |
| `arrange()` | reorder rows |

: {tbl-colwidths="[22,78]"}

### select()

`select()` keeps the columns you name, in the order you name them. Note that the names go **without** quotation marks.

```{r}
raw |>
  select(F2001_A, F2002, F3020_R) |>
  head(n = 3)
```

You can also drop columns with a minus sign, and select ranges with a colon. `starts_with()` is one of several helpers that pick columns by pattern:

```{r}
raw |>
  select(starts_with("F3018_")) |>
  names()
```

Those are the eight party ratings. `starts_with()` will earn its keep repeatedly in this data set, because CSES names its batteries systematically.

### rename()

`rename()` changes names, in the form `new = old`.

```{r}
raw |>
  select(F2001_A, F3020_R) |>
  rename(age = F2001_A, lr_self = F3020_R) |>
  head(n = 3)
```

Note which side is which. It is `new = old`, the same direction as `<-`: the thing being created is on the left.

### filter()

`filter()` keeps rows where a condition is `TRUE`.

```{r}
raw |>
  filter(F2001_A < 30) |>
  nrow()
```

`nrow()` counts rows. Several conditions can be given, separated by commas, and all of them must hold:

```{r}
raw |>
  filter(F2001_A < 30, F2002 == 1) |>
  nrow()
```

Here is a real advantage over base R. Last week we needed `which()` to stop missing values from producing phantom rows. `filter()` does that for you: a row whose condition is `NA` is dropped, not kept as a row of `NA`s.

> That convenience has a cost worth knowing about. `filter()` **silently** removes rows where the condition cannot be evaluated. If you filter on a variable with a lot of missing data, you lose those respondents without being told. Check with `nrow()` before and after when it matters.

### mutate()

`mutate()` adds columns. Everything else so far has been about choosing what you already have; this is where new things get made.

```{r}
raw |>
  select(F3018_A, F3018_B) |>
  mutate(gap = F3018_A - F3018_B) |>
  head(n = 3)
```

`F3018_A` is the rating of the Social Democrats and `F3018_B` the rating of the Sweden Democrats, each 0 to 10. Their difference is a new quantity: how much more one respondent likes one than the other.

The first three rows look sensible. Now ask what the whole column looks like:

```{r}
raw |>
  mutate(gap = F3018_A - F3018_B) |>
  pull(gap) |>
  range()
```

`pull()` takes a single column out of a data frame as a plain vector, which is what `range()` wants.

::: {.discussion}
**A question.** Both ratings run from 0 to 10, so the difference between them has hard limits of −10 and +10. The range above is −99 to 99.

Where do those values come from, and how many rows are affected? Note that the first three rows gave no hint of the problem, and that `mutate()` reported no error.

What does that suggest about the order in which cleaning and calculating should be done?
:::

There are `r sum(abs(raw$F3018_A - raw$F3018_B) > 10)` such rows. This is the same lesson as last week from the other direction: derive a quantity from uncleaned variables and the contamination spreads into the new variable, where it is harder to spot because the new variable has no codebook.

Columns made inside one `mutate()` can be used by later ones in the same call, which lets a calculation be written in readable steps rather than in one nested expression.

### arrange()

`arrange()` sorts. `desc()` reverses the direction.

```{r}
raw |>
  select(F2001_A, F3020_R) |>
  arrange(desc(F2001_A)) |>
  head(n = 3)
```

Missing values always go to the end, whichever direction you sort in.

## Recoding with case_when()

Last week we recoded by creating an all-missing variable and filling it in with square brackets. `case_when()` does the same job in one expression.

It takes a series of `condition ~ value` pairs, checks them **in order**, and gives each row the value from the first condition that is `TRUE`. The `~` separates the test from the result.

```{r}
recoded <- raw |>
  mutate(
    agegroup = case_when(
      F2001_A < 35 ~ "Under 35",
      F2001_A < 65 ~ "35 to 64",
      F2001_A >= 65 ~ "65 and over"
    )
  )

table(recoded$agegroup, useNA = "ifany")
```

Because the conditions are checked in order, the second one does not need to say `F2001_A >= 35` — anything under 35 has already been dealt with. That is convenient, and it is also a trap: reorder the lines and the answer changes.

Anything matching no condition becomes `NA`. That is the same safe default we built by hand last week, and here it comes for free.

To count something rather than label it, `if_else()` handles the two-way case: condition, value if true, value if false.

```{r}
recoded <- raw |>
  mutate(retired = if_else(F2006 == 7, 1, 0))
```

## Doing the same thing to many columns

This is where the tidyverse stops being a convenience and starts being necessary.

Our data has a seven-item trust battery, an eight-party rating battery, an eight-party left–right battery and an eight-leader battery. That is thirty-one columns needing the same treatment: set the missing codes to `NA`, and for some of them reverse the scale. Writing that out one column at a time is thirty-one chances to make a typo, and typos of this kind do not produce errors.

`across()` applies a function to a set of columns. It goes inside `mutate()` and takes two arguments: which columns, and what to do to them.

```{r}
trust <- raw |>
  select(starts_with("F3007_")) |>
  mutate(
    across(
      everything(),
      \(x) if_else(x %in% c(7, 8, 9), NA, x)
    )
  )

summary(trust$F3007_1)
```

Three things in there are new.

`everything()` means all the columns currently selected — here, the seven trust items.

`\(x)` creates a small unnamed function on the spot. Read it as "given a column, which I will call `x`, do the following to it". The backslash is shorthand; you may also see it written `function(x)`, which means exactly the same. `across()` calls it once per column, and the value of `x` is that column.

`if_else(x %in% c(7, 8, 9), NA, x)` says: where the value is one of the missing codes, put `NA`; otherwise leave it alone.

Reversing all seven at once is the same shape:

```{r}
trust <- trust |>
  mutate(across(everything(), \(x) 5 - x))

summary(trust$F3007_1)
```

The scale now runs 1 to 4 with 4 meaning most trust, for all seven items, from one line.

> `across()` is powerful in the way a chainsaw is powerful. It applies your function to every column you select, whether or not that makes sense for each of them. Reversing with `5 - x` is right for a four-point scale and wrong for a 0-to-10 scale, and nothing will stop you doing it to both. Select deliberately, and check the ranges afterwards.

## Grouping and summarising

The second thing the tidyverse does much better than base R: calculating something separately for each group.

`summarise()` collapses a data frame to a single row.

```{r}
raw |>
  filter(F3020_R <= 10) |>
  summarise(
    n = n(),
    mean_lr = mean(F3020_R),
    sd_lr = sd(F3020_R)
  )
```

`n()` counts the rows in the current group. The `filter()` at the top removes the missing codes, which on this 0-to-10 variable are 95 and above.

`group_by()` splits the data first, so that `summarise()` runs once per group.

```{r}
raw |>
  filter(F3020_R <= 10, F2002 %in% c(0, 1)) |>
  group_by(F2002) |>
  summarise(
    n = n(),
    mean_lr = mean(F3020_R)
  )
```

`F2002` is gender, coded 0 for men and 1 for women. Two groups, so two rows.

The reason this matters is that the number of groups does not change the code. Here is the same calculation across all eight parties:

```{r}
raw |>
  filter(F3020_R <= 10, F3011_LH_PL < 999000) |>
  group_by(F3011_LH_PL) |>
  summarise(
    n = n(),
    mean_lr = round(mean(F3020_R), 2)
  ) |>
  arrange(mean_lr)
```

Eight rows, sorted from left to right. The party codes are still raw numbers — we fix that below — but the ordering is already recognisable if you know Swedish politics: the Left Party at one end, the Sweden Democrats and Moderates at the other.

> `group_by()` stays switched on. After a `summarise()`, the result may still be grouped, and the next thing you do will happen within groups rather than overall. When you are finished with the grouping, end with `ungroup()`.

## Joining

Often the information you need is spread over two files. `F2018` holds a county code; the county's name and which of the three historical lands it belongs to live in a separate lookup table.

```{r}
counties <- readr::read_csv("data/counties.csv", show_col_types = FALSE)
head(counties, n = 3)
```

`readr::read_csv()` rather than `read.csv()`. The two colons mean "the `read_csv` function from the `readr` package", which lets you use one function from a package without loading the whole thing. `show_col_types = FALSE` suppresses the message it otherwise prints about what type it decided each column is. The reason for preferring `read_csv()` here appears in a moment.

`left_join()` adds columns from the second table to the first, matching rows on a shared column. The `by` argument says which columns to match on, in the form `"name in the first" = "name in the second"`.

```{r}
joined <- raw |>
  select(F2018, F3020_R) |>
  left_join(counties[, c("code", "county", "land")], by = c("F2018" = "code"))

head(joined, n = 3)
```

Only three columns of `counties` are taken, because a join brings across everything you do not exclude, and unused columns accumulate quickly.

Every row of `raw` is kept — that is what makes it a *left* join. Rows with no match get `NA` in the new columns.

Which makes the check obvious and necessary:

```{r}
sum(is.na(joined$county))
```

`r sum(is.na(joined$county))` rows failed to match, out of `r format(nrow(joined), big.mark = " ")`. Those are respondents with no county recorded. That is a small enough number to ignore — but you only know that because you counted.

### When a join goes wrong

Here is why that check is not a formality, and why the `readr::read_csv()` above was not fussiness.

The county names contain å, ä and ö. Read the same file two ways and look at what R records about the text:

```{r}
counties_base <- read.csv("data/counties.csv")

unique(Encoding(counties_base$county))
unique(Encoding(counties$county))
```

`Encoding()` reports what R believes each string to be. `read.csv()` says `unknown` for all of them, meaning "whatever this machine's native encoding is". `readr::read_csv()` gives two answers: `UTF-8` for the names containing å, ä or ö, and `unknown` for the plain ones — which is right, because text made only of ordinary English letters is identical in every encoding and needs no mark.

On this machine the native encoding *is* UTF-8, so the two agree and everything works. On a machine where it is not — and Windows has historically used something else — `unknown` and `UTF-8` are different, and R will not treat the two strings as equal.

**Identical bytes. Different answer, depending on the computer.**

This is not hypothetical. While these materials were being prepared, exactly this join was run with the county table read by `read.csv()` in a session whose locale was not UTF-8. It matched `r 2845 - 1695` rows of `r format(2845, big.mark = " ")` and silently turned the other 1,695 into `NA`. There was no error and no warning. Every county whose name is plain English matched — Stockholm, Uppsala, Kalmar, Gotland — and every county with a diacritic did not. The only visible symptom was a category with a count of zero.

Had the next step been a table of results, there would have been a table of results: perfectly plausible, and computed on the counties that happen not to have diacritics in their names.

Three habits, in order of how much they buy you:

- **Count the rows after every join.** One line. It catches this and every other reason a join can fail.
- **Read text data with `readr::read_csv()`**, which records the encoding rather than guessing.
- **Pin the locale** at the top of any script that touches non-English text, so the behaviour does not depend on whose computer it runs on:

```{r}
#| eval: false
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

> The general point outlives this particular bug. An error that stops your script is cheap — you find it immediately and you fix it. An error that returns a *wrong answer* costs you whatever you build on top of it. Text comparison is one of the places the second kind lives, and a join is where it does the most damage.

Now the county information is attached, grouping by it works like any other variable:

```{r}
joined |>
  filter(F3020_R <= 10, !is.na(land)) |>
  group_by(land) |>
  summarise(
    n = n(),
    mean_lr = round(mean(F3020_R), 2)
  )
```

`!is.na(land)` keeps the rows where `land` is *not* missing — `!` means "not".

## Building the analysis data set

Everything so far has been on fragments. The real job is to do it to all `r ncol(readRDS("data/swe.rds"))` variables, and that is what the script `build_swe.R` does. It ships with the course data, and you should read it — it is the worked example for this whole session.

It is too long to print here, but it is built from four moves, all of which you have now seen.

**One.** Two small helper functions, `set_na()` and `reverse()`, so that the same operation is written once rather than eighty times.

```{r}
set_na <- function(x, codes) {
  x[x %in% codes] <- NA
  x
}

reverse <- function(x, lo, hi) {
  (hi + lo) - x
}
```

A function you write yourself is defined with `function(arguments)` followed by the body in curly brackets. The last expression in the body is what comes back out.

**Two.** Each variable is cleaned and given a short, readable name.

```{r}
#| eval: false
swe$polint <- set_na(raw$F3001, c(7, 8, 9))
swe$polint <- reverse(swe$polint, 1, 4)
```

**Three.** Batteries are handled as a group, so that all seven trust items — or all eight party ratings — are treated identically by construction rather than by care.

**Four.** Every recoded variable is checked against the range it is not allowed to leave, and the script stops if any of them does. This is the same `stopifnot()` idea as last week, applied to `r nrow(read.csv("data/codebook.csv"))` variables at once instead of one.

The result is `swe.rds`, which everything from session 4 onwards uses:

```{r}
swe <- readRDS("data/swe.rds")
dim(swe)
```

`r nrow(swe)` respondents and `r ncol(swe)` variables, with short names, missing values marked as missing, scales pointing the same way, and factors in a sensible order. A codebook describing every variable is in `codebook.csv`.

```{r}
swe |>
  select(vote, lr_self, satdem, tr_parl) |>
  head(n = 4)
```

::: {.discussion}
**A question.** The script does all of this from the raw file every time it runs, rather than someone fixing the data once by hand and saving the result.

Given how much longer it takes to write a script than to edit a few values, why is it done this way?
:::

## Saving

`saveRDS()` writes an R object, and `write.csv()` writes plain text. The `.rds` version is smaller and keeps factors as factors; the CSV can be opened by anything.

```{r}
#| eval: false
saveRDS(swe, "data/swe.rds")
write.csv(swe, "data/swe.csv", row.names = FALSE)
```

> If your data contains å, ä, ö or any other letter outside plain English, say so explicitly when you save. `readr::write_excel_csv()` writes UTF-8 and marks it, which avoids the silent comparison failure described at the end of last week's chapter.

::: {.swirl}
**In the seminar.** `swirl` lesson 3, *Data and Data Management II*. The pipe, the five verbs, `group_by()` and `summarise()`, `across()`, and a join. Around 30 minutes.
:::
