# The logic of hypothesis testing

```{r}
#| label: setup-06
#| include: false
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
options(width = 78)

# Fixed number of decimal places, so that a trailing zero is not dropped.
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

Everything up to now has described the data in front of us. From here on we want to say something about the world the data came from — and that is a different and much harder claim.

This chapter has no new method in it. It is the reasoning that every method in the rest of the course depends on, and it is worth slowing down for, because almost every mistake people make with statistics is a mistake about this rather than about arithmetic.

```{r}
library(ggplot2)
library(dplyr)

swe <- readRDS("data/swe.rds")
turnout <- readr::read_csv("data/turnout.csv", show_col_types = FALSE)
```

## A sample is not a population

The **population** is the group you want to know about: everyone entitled to vote in Sweden in 2022. The **sample** is the `r nrow(swe)` people who actually answered.

You want to talk about the first and you only have the second. Everything in this chapter is about what that costs.

Start with something we can check, because we know the right answer. Our survey reports how many people voted. So does the electoral authority.

```{r}
turnout |>
  filter(study == "SWE_2022") |>
  select(country, n, reported, official, gap)
```

`r turnout$reported[turnout$study == "SWE_2022"]`% against `r turnout$official[turnout$study == "SWE_2022"]`%. The survey is out by `r turnout$gap[turnout$study == "SWE_2022"]` percentage points.

Now the important part. That could be bad luck — this particular sample happening to catch more voters than average. If so, another study would miss in the other direction. CSES Module 6 contains 18 election studies in 17 countries, run by different organisations using different methods, so we can look.

```{r}
#| label: fig-turnout
#| fig-height: 4.4
turnout |>
  mutate(
    label = paste(country, year),
    label = reorder(label, gap)
  ) |>
  ggplot(aes(x = gap, y = label)) +
  geom_col(fill = "grey60") +
  geom_vline(xintercept = 0) +
  labs(
    x = "Reported turnout minus official turnout (percentage points)",
    y = NULL
  ) +
  theme_bw()
```

`geom_col()` draws bars whose length is a value in the data, as against `geom_bar()`, which counts rows. Each bar is one study, labelled with `paste()` as country and year: Portugal has two, and labelled by country alone its two bars would be stacked into one, as long as both gaps added together. `reorder()` sorts one variable by another, so the studies come out in order of the size of the gap rather than alphabetically. `geom_vline()` draws a vertical line, with `xintercept` saying where on the horizontal axis it sits. `y = NULL` removes the axis title, because "label" over a list of countries and years carries nothing.

Every single one is positive.

```{r}
turnout |>
  summarise(
    studies = n(),
    over_reporting = sum(gap > 0),
    smallest = min(gap),
    largest = max(gap)
  )
```

Eighteen studies, eighteen over-reports, from `r min(turnout$gap)` to `r max(turnout$gap)` percentage points. Nothing that goes the same way eighteen times out of eighteen is bad luck.

This is the distinction the whole chapter turns on.

**Random error** is the sample missing the population by chance. It is unavoidable, it goes both ways, it shrinks as the sample grows, and — this is the useful part — **it can be measured**. Most of this chapter is about measuring it.

**Systematic error** is the sample missing the population in a consistent direction. Here, people who do not vote are less likely to answer a survey about an election, and some of those who do answer say they voted when they did not. It does not go both ways, **it does not shrink as the sample grows**, and no statistical method in this course will detect it or fix it.

> A bigger sample buys you less random error and nothing whatever against systematic error. A survey of ten million people, all of them recruited in the same biased way, is wrong by exactly as much as a survey of a thousand — it is just more confident about it.

### What a random sample is

The methods in the rest of this course all assume a **random sample**: every member of the population has a known, non-zero chance of being selected.

The Swedish study comes close. Statistics Sweden drew names from the population register, which is about as good as sampling gets. What it could not control was who answered — and that is where the turnout gap comes from. A perfect sampling frame does not save you if response is selective.

::: {.discussion}
**A question.** The survey also reports `r round(100 * prop.table(table(swe$vote))[["SD"]], 1)`% voting for the Sweden Democrats, where the official result recorded by the Swedish electoral authority was 20.5%.

Is that the same problem as the turnout gap, or a different one? Which of the two would a larger sample help with?
:::

> When your cases are countries rather than people, you never have a random sample — there is only one world and you have all of it, or the part of it that publishes data. Social scientists routinely run these methods on country data anyway and treat the result as if the assumption held. You should know that this is what is happening.

## Measuring variation

To measure random error we first need to measure spread. How far apart are the values in a variable?

The natural starting point is how far each case sits from the mean.

```{r}
lr <- swe$lr_self[!is.na(swe$lr_self)]

deviations <- lr - mean(lr)
round(sum(deviations), 10)
```

Zero. Not approximately — exactly, up to rounding. That is what a mean *is*: the point where the distances above and below balance. So the sum of deviations is useless as a measure of spread; it is zero for every variable that ever existed.

The fix is to square the deviations first, which makes them all positive, then average them. That average is the **variance**.

```{r}
squared <- deviations^2
sum(squared) / (length(lr) - 1)
var(lr)
```

Squaring solved one problem and created another: the variance is in squared units. For a 0-to-10 scale it is in "scale points squared", which means nothing. Take the square root and you are back on the original scale. That is the **standard deviation**.

```{r}
sd(lr)
```

Roughly, the standard deviation is the typical distance of a case from the mean. Here it is about `r round(sd(swe$lr_self, na.rm = TRUE), 1)` points on a scale running from 0 to 10.

In formulas, with $\bar{x}$ for the mean and $n$ for the number of cases:

$$s^2 = \frac{\sum(x_i - \bar{x})^2}{n - 1} \qquad s = \sqrt{s^2}$$

### Why $n - 1$

Notice that we divided by `length(lr) - 1` rather than by `length(lr)`, and that `var()` agreed.

The reason is that we are not describing these `r length(swe$lr_self[!is.na(swe$lr_self)])` people. We are using them to estimate the spread in the population. And a sample is systematically *less* spread out than the population it came from, because the rare extreme cases are, by definition, rarely sampled.

So the obvious calculation would be biased downwards, every time. Dividing by $n-1$ instead of $n$ makes the result slightly larger and corrects it. With thousands of cases the difference is invisible; with twelve it matters.

## What happens when you sample repeatedly {#sec-repeated}

Here is the central idea, and R lets us watch it happen rather than take it on trust.

Suppose our `r length(lr)` respondents were the whole population, and their mean left-right position, `r round(mean(lr), 2)`, were the true value. Now draw a small sample of 50 from them and take its mean.

```{r}
set.seed(2026)

mean(sample(lr, size = 50))
```

`sample()` draws at random from a vector; `size` says how many. `set.seed()` fixes the random number generator so that you get the same "random" draw as this page did — without it your number will differ, which is the point.

That sample missed. Draw another and it will miss differently. The question is: **by how much, typically?**

We can just do it. `replicate()` runs an expression many times and collects the results.

```{r}
sample_means <- replicate(2000, mean(sample(lr, size = 50)))

length(sample_means)
```

Two thousand samples, two thousand means. What do they look like?

```{r}
#| label: fig-sampling
#| fig-height: 3.6
ggplot(data.frame(sample_means), aes(x = sample_means)) +
  geom_histogram(binwidth = 0.1, fill = "grey60") +
  geom_vline(xintercept = mean(lr), linewidth = 1) +
  labs(
    x = "Mean of a sample of 50",
    y = "Number of samples"
  ) +
  theme_bw()
```

Three things to notice, and all of them matter.

**It is centred on the truth.** The vertical line — drawn by `geom_vline()` again, with `linewidth` making it thicker than the default — is the population mean. The samples scatter around it rather than drifting off somewhere else.

```{r}
mean(sample_means)
mean(lr)
```

**It is bell-shaped**, even though the variable itself is not — `lr_self` is a lumpy 0-to-10 scale with a spike at 5. The distribution of *sample means* comes out smooth and symmetric anyway. This happens almost regardless of the shape of the original variable, which is the single most useful fact in statistics.

**Its spread is measurable.** That is the number we actually want:

```{r}
sd(sample_means)
```

That is how far a sample of 50 typically misses by. It has a name: the **standard error**.

### The point of all that

We simulated it because we could — we had all `r length(lr)` cases and could pretend they were a population. In real research you have one sample and no way to draw another.

The remarkable thing is that you do not need to. The standard error can be calculated from a single sample:

$$SE = \frac{s}{\sqrt{n}}$$

The standard deviation, divided by the square root of the sample size. Compare it with what the simulation produced:

```{r}
sd(lr) / sqrt(50)
sd(sample_means)
```

`r fmt(sd(lr)/sqrt(50), 3)` against `r fmt(sd(sample_means), 3)`. The formula tells you, from one sample, what two thousand samples would have shown you.

Two consequences follow from the $\sqrt{n}$ on the bottom.

**More data means less random error** — the standard error shrinks as the sample grows.

**But slowly.** To halve the standard error you need **four times** the data. This is why the difference between a sample of 1,000 and a sample of 2,000 is much smaller than people expect, and why the useful question about a survey is almost never "is it big enough".

::: {.discussion}
**A question.** Our full sample has `r length(lr)` valid answers on `lr_self`, not 50.

Work out the standard error of the mean for the full sample. Then ask what would have to be true for that number to be a good description of how far our estimate is from the truth about Sweden — and whether the turnout table at the top of this chapter gives you any confidence that it is.
:::

## The 95% rule {#sec-95}

Because sampling distributions come out bell-shaped, one fact about the normal distribution does most of the practical work.

In a normal distribution, about **95% of the values lie within about two standard deviations of the mean**. More precisely, 1.96.

```{r}
pnorm(1.96)
pnorm(1.96, lower.tail = FALSE)
```

`pnorm()` gives the proportion of a normal distribution below a given number of standard deviations from the mean. Below 1.96 lies `r round(100 * pnorm(1.96), 1)`% of it; `lower.tail = FALSE` asks for the other end instead, giving the `r round(100 * pnorm(1.96, lower.tail = FALSE), 1)`% above. Two tails of 2.5% each leaves 95% in the middle.

Applied to a sampling distribution, that says: about 95% of samples produce an estimate within 1.96 standard errors of the truth.

## Confidence intervals

Turn that round and you have the most useful thing in this chapter.

```{r}
estimate <- mean(lr)
se <- sd(lr) / sqrt(length(lr))

lower <- estimate - 1.96 * se
upper <- estimate + 1.96 * se

round(c(estimate, lower, upper), 3)
```

The mean is `r round(mean(lr), 2)`, with a 95% confidence interval from `r round(mean(lr) - 1.96 * sd(lr)/sqrt(length(lr)), 2)` to `r round(mean(lr) + 1.96 * sd(lr)/sqrt(length(lr)), 2)`.

Read it as: **a range of values for the population figure that this sample is consistent with.** Values inside the interval are ones the data cannot rule out. Values outside it are ones the data argues against.

> **What a confidence interval does not mean.** It is tempting to say "there is a 95% probability that the true value is in this interval". That is the standard misstatement, and it is worth knowing why.
>
> The population value is a fixed number. It is either inside this interval or it is not; there is no probability about it. The 95% describes **the procedure**, not this particular interval: if you drew sample after sample and built an interval each time, about 95% of those intervals would contain the true value. You never learn whether yours is one of them.
>
> In practice people read a confidence interval as a range of plausible values, and that is a reasonable working habit. But when you write it up, write it as a range, not as a probability.

And note what it is silent about. The interval says nothing about the `r turnout$gap[turnout$study == "SWE_2022"]`-point turnout gap. It measures random error only. **A narrow confidence interval on a biased estimate is a precise wrong answer.**

## Hypothesis tests

The last piece. Most research questions are not "what is the value" but "is there anything there at all" — does this differ between groups, is this related to that.

Such questions get turned into a pair of hypotheses:

- The **null hypothesis**: there is nothing there. No difference, no association — the population value is the one that "nothing" would give.
- The **alternative hypothesis**: there is something there. The population value is not that.

A test asks one thing: **how far is the sample from what the null hypothesis says, and is that further than sampling noise would carry it?** The rest of this section builds that up one step at a time, using a question worth asking.

### A null hypothesis worth testing

The left-right scale runs from 0 to 10, and 5 is its middle. A claim you will hear is that Swedish voters are, on average, exactly in the centre. That gives a pair of hypotheses:

- Null: the mean left-right position of Swedish voters is **5**.
- Alternative: it is **not 5** — it is somewhere else, in either direction.

The sample mean is `r round(estimate, 2)`, not 5. But a sample mean is never exactly the population value; @sec-repeated showed sample means scattering around the truth. The question is whether `r round(estimate, 2)` is the kind of miss a sample makes by chance when the truth is 5.

### Step 1: how far is the sample from the null?

```{r}
null_value <- 5
distance <- estimate - null_value
distance
```

The sample is `r round(distance, 2)` points from where the null hypothesis puts the population. On its own that number cannot answer the question. Is `r round(distance, 2)` points a lot? It depends on how much a sample mean moves about from sample to sample — and that is exactly what the standard error measures.

### Step 2: measure the distance in standard errors

The standard error is the yardstick. @sec-repeated showed what it is: the spread of the means you would get from sample after sample. So divide the distance by it:

```{r}
z <- distance / se
z
```

The sample mean sits `r round(z, 2)` **standard errors** from the null value. That is the **test statistic**:

$$z = \frac{\text{estimate} - \text{null value}}{\text{standard error}}$$

This is the signal-to-noise ratio from the start of the section, made concrete. The signal is the distance between what we found and what the null predicts. The noise is the standard error — how far a sample mean typically wanders by chance alone.

### Step 3: where does that land if the null is true?

Now suppose the null hypothesis were true. Then the population mean would be 5, and the means of repeated samples would pile up around 5 in the bell shape you saw in @sec-repeated, with a spread of one standard error. Measured in standard errors, that is a normal distribution centred on 0.

The test statistic is **a location in that distribution**: how far out our sample sits among all the samples a world with no effect would produce.

```{r}
#| label: fig-null
#| echo: false
#| fig-cap: "Where sample means would fall if the null hypothesis were true. The bottom axis measures the distance from the null value in standard errors; the top axis gives the same positions as mean left-right placements, with the null value of 5 at the centre. The shaded tails are the 5% of samples furthest from it, beyond ±1.96. The vertical line is our sample."
#| fig-height: 3.2
curve_df <- data.frame(x = seq(-5, 5, by = 0.01))
curve_df$y <- dnorm(curve_df$x)
ggplot(curve_df, aes(x = x, y = y)) +
  geom_area(data = subset(curve_df, x <= -1.96), fill = "grey75") +
  geom_area(data = subset(curve_df, x >= 1.96), fill = "grey75") +
  geom_line() +
  geom_vline(xintercept = z, linewidth = 1) +
  annotate("text", x = z, y = 0.3, label = "our sample",
           hjust = 1.1, size = 4) +
  scale_x_continuous(
    sec.axis = sec_axis(\(x) null_value + x * se,
                        name = "Mean left-right position",
                        breaks = seq(4.8, 5.3, by = 0.1))
  ) +
  labs(
    x = "Distance from the null value, in standard errors",
    y = NULL
  ) +
  theme_bw() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

At `r round(z, 1)` standard errors, our sample is past the point where the curve has practically run out. Samples that far from 5 almost never happen when the truth is 5.

### Step 4: from a location to a probability

The **p-value** turns that location into a number: the share of the null distribution that lies **at least as far from the null value as our sample**, in either direction.

```{r}
z_size <- abs(z)
beyond <- pnorm(z_size, lower.tail = FALSE)
p <- 2 * beyond
p
```

`abs()` gives the size of a number without its sign, because a sample `r round(z, 1)` standard errors *below* 5 would be just as much evidence against the null as one `r round(z, 1)` above. `pnorm()` with `lower.tail = FALSE`, from @sec-95, gives the share of the normal distribution beyond that point in one tail. The alternative was "not 5, in either direction", so both tails count, and the share is doubled.

The printed `2.78e-05` is how R writes `r format(p, scientific = FALSE, digits = 3)`: the number after `e-` says how many places to move the decimal point to the left. If the average Swede really sat at exactly 5, a sample as far from it as ours would turn up about once in `r format(round(1 / p, -3), big.mark = ",")` samples. That is the p-value: **the probability of a result at least this far from the null, if the null were true.**

So the no-effect story is a poor explanation of these data. The conventional threshold is 0.05, and this is far below it.

> **Read that definition again**, because the direction catches people out. The p-value is the probability of the *data* given no effect. It is **not** the probability that there is no effect, and it is not the probability that you are wrong. It cannot be, because it assumes the null hypothesis to compute the number, and you cannot use an assumption as evidence about itself.

### The same distance, different evidence

Here is why the standard error matters so much. Suppose the same mean, `r round(estimate, 2)`, had come from only 100 respondents rather than `r format(length(lr), big.mark = ",")`. The distance from 5 is identical. The standard error is not.

```{r}
se_100 <- sd(lr) / sqrt(100)
z_100 <- distance / se_100
p_100 <- 2 * pnorm(abs(z_100), lower.tail = FALSE)
round(c(se_100, z_100, p_100), 3)
```

With 100 respondents the standard error is `r round(se_100, 2)` rather than `r round(se, 3)`, the same `r round(distance, 2)` points is only `r round(z_100, 2)` standard errors, and the p-value is `r fmt(p_100)`.

```{r}
#| label: fig-null-100
#| echo: false
#| fig-cap: "The same distance from 5 in a sample of 100. The top axis is wider than before because the standard error is larger. The shaded area is the p-value: the share of samples at least this far from the null, in either direction, if the null were true."
#| fig-height: 3.2
ggplot(curve_df, aes(x = x, y = y)) +
  geom_area(data = subset(curve_df, x <= -z_100), fill = "grey75") +
  geom_area(data = subset(curve_df, x >= z_100), fill = "grey75") +
  geom_line() +
  geom_vline(xintercept = z_100, linewidth = 1) +
  scale_x_continuous(
    sec.axis = sec_axis(\(x) null_value + x * se_100,
                        name = "Mean left-right position",
                        breaks = seq(3.75, 6.25, by = 0.25))
  ) +
  labs(
    x = "Distance from the null value, in standard errors",
    y = NULL
  ) +
  theme_bw() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

Now the p-value is visible as an area: `r round(100 * p_100)`% of samples would land at least that far from 5 even if 5 were the truth. A sample like that is entirely ordinary in a world with no effect, so it is no evidence against the null.

The distance was the same both times. **What changed was the standard error**, and with it the location of the sample in the null distribution. That is how the size of a sample enters a test: through the square root of *n* in the standard error.

### The rule of 1.96, again

@sec-95 showed that 95% of a normal distribution lies within 1.96 standard deviations of its centre. So a sample more than 1.96 standard errors from the null value lands in the outer 5% — and that is exactly when the p-value falls below 0.05.

It is also exactly when the 95% confidence interval leaves out the null value, because the interval is the estimate plus and minus 1.96 standard errors. Confidence intervals and p-values are two views of the same calculation. Our interval runs from `r round(estimate - 1.96 * se, 2)` to `r round(estimate + 1.96 * se, 2)`: it does not include 5, and the p-value is below 0.05.

### A test is only as interesting as its null

The null hypothesis of 5 was worth testing: it is a claim people make. Testing whether the mean differs from 0 would give an even smaller p-value and mean nothing at all, because there was never any reason to think the average Swede sits at the far left of the scale. **A test is only interesting if the null hypothesis was worth entertaining.** Much published research reports p-values for comparisons nobody doubted.

The next three sessions are all applications of this one idea. Session 7 tests whether two categorical variables are related, session 8 whether two means differ, session 9 whether two continuous variables move together. Each has its own estimate and its own standard error, and R does the steps above in one function call, but the logic is always these four steps: a distance from the null, measured in standard errors, located in the distribution the null predicts, and read off as a probability.

> With small samples the functions you will meet, such as `t.test()` in session 8, use the *t* distribution instead of the normal — a slightly wider bell that allows for the standard error itself being estimated from a small sample. With thousands of cases the two are indistinguishable, and `t.test()` gives the same answer as the steps above.

::: {.reporting}
**How to write it up.**

> Swedish respondents place themselves close to the centre of the left-right scale (mean = `r round(estimate, 2)`, 95% CI [`r round(estimate - 1.96 * se, 2)`, `r round(estimate + 1.96 * se, 2)`], *n* = `r format(length(lr), big.mark = ",")`), but slightly to the right of its midpoint of 5 (*z* = `r round(z, 2)`, *p* < 0.001).

Report the estimate, the interval, and the number of cases the estimate rests on. For a test, add the test statistic and the p-value — and say what the null value was. Mistakes to avoid:

- **Reporting a p-value without the estimate.** "The difference was significant (p < .05)" does not say how big it was, which is what the reader needs.
- **Reporting more decimal places than the data supports.** A left-right scale with eleven whole-number points does not measure anything to three decimals.
- **Calling a result "significant" in prose without the word "statistically".** In ordinary English significant means important, and a statistically significant finding can be far too small to matter.
- **Treating p just above 0.05 as no effect and just below as an effect.** 0.049 and 0.051 are the same evidence. The threshold is a convention, not a discovery.
- **Forgetting the *n***, especially after missing values have quietly removed part of the sample.
:::

::: {.swirl}
**In the seminar.** `swirl` lesson **06 Hypothesis Testing**. Variation, drawing repeated samples and watching the sampling distribution appear, the standard error, and confidence intervals. Around 30 minutes.
:::
