# Regression: presenting results

```{r}
#| label: setup-12
#| include: false
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
options(width = 78)
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

You have a model. Now somebody else has to read it.

The raw output of `summary()` is for you, while you are working. It is not for a reader: it prints things nobody needs, uses variable names instead of words, and gives eight significant digits for a quantity known to one. This session is about the difference.

```{r}
library(ggplot2)
library(dplyr)
library(psych)
library(modelsummary)
library(ggeffects)

swe <- readRDS("data/swe.rds")

trust_items <- c(
  "tr_parl", "tr_govt", "tr_court",
  "tr_sci", "tr_party", "tr_media"
)
swe$trust <- fa(swe[, trust_items], nfactors = 1, fm = "ml")$scores[, 1]
```

Three models, of increasing size — which is itself the standard way to present regression results.

```{r}
model1 <- lm(demlevel ~ trust, data = swe)
model2 <- lm(demlevel ~ trust + bloc, data = swe)
model3 <- lm(demlevel ~ trust + bloc + age + educ, data = swe)
```

## Regression tables

`modelsummary()` takes a list of models and puts them side by side.

```{r}
modelsummary(list(model1, model2, model3))
```

Better than three blocks of `summary()` output, but not yet publishable. Three things are wrong: the variable names are code, there is more in the bottom panel than a reader needs, and the columns are unlabelled.

```{r}
labels <- c(
  "(Intercept)" = "Intercept",
  "trust" = "Trust in institutions",
  "blocRight" = "Voted for the right bloc",
  "age" = "Age (years)",
  "educ" = "Education (ISCED level)"
)

modelsummary(
  list("Trust only" = model1,
       "Plus bloc" = model2,
       "Full model" = model3),
  coef_map = labels,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  stars = TRUE,
  title = "Evaluations of Swedish democracy (0-10)"
)
```

`coef_map` renames the coefficients and, importantly, also **sets their order** — anything left out of the map is dropped from the table. Naming the list of models gives the columns headings. `gof_map` picks which goodness-of-fit rows to keep. `stars = TRUE` adds significance stars and a footnote explaining them.

The number in brackets under each coefficient is its standard error. Some journals want confidence intervals instead:

```{r}
modelsummary(
  list("Full model" = model3),
  coef_map = labels,
  gof_map = c("nobs", "r.squared"),
  estimate = "{estimate}",
  statistic = "conf.int"
)
```

`statistic` controls what goes underneath. `"conf.int"` gives the 95% interval, which is more informative than a standard error because the reader does not have to do the arithmetic.

> **Stars or intervals?** Stars are conventional and compact, and they encourage exactly the habit session 6 warned against — reading a table for which rows have asterisks. A confidence interval shows the size of the effect and its uncertainty at once. Give intervals where the journal permits it.

To save a table to a file, give `output` a filename. The format follows the extension.

```{r}
#| eval: false
modelsummary(
  list("Full model" = model3),
  coef_map = labels,
  output = "tables/democracy_model.docx"
)
```

`.docx`, `.html`, `.tex` and `.md` all work. Writing straight into the document you are drafting means the numbers cannot be mistyped on the way — the same argument as inline code in these materials.

> You will also meet **stargazer**, which was the standard for years and appears in a great deal of published code. It still works for `lm()`, but it has not kept up with newer model types. Use `modelsummary` for new work; recognise `stargazer` when you read someone else's.

## Coefficient plots

A table is precise. A plot is quicker to read, and for a model with many predictors it is much quicker.

```{r}
#| label: fig-coef
#| fig-height: 3.2
modelplot(
  model3,
  coef_map = rev(labels[-1]),
  conf_level = 0.95
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Coefficient (points on the 0-10 scale)",
    y = NULL
  ) +
  theme_bw()
```

`modelplot()` comes from the same package and returns an ordinary ggplot, so everything from session 5 applies. `labels[-1]` drops the intercept, which is not a comparable quantity and would compress everything else; `rev()` reverses the order, because a plot builds upwards from the bottom while a table reads downwards.

The dashed line at zero is doing real work. **Any interval crossing it is a coefficient we cannot distinguish from no relationship.**

::: {.discussion}
**A question.** On that plot, trust has a long bar far from zero and age has a very short bar sitting on it.

A reader might conclude that trust matters and age does not. What else would you need to know before agreeing — and what does the fact that the two variables are on completely different scales do to the comparison?
:::

That question points at a real limitation. The coefficients are in different units — one standard deviation of trust, one year of age, one ISCED level — so their lengths on this plot are not comparable. One fix is to standardise the predictors so that every coefficient is "per standard deviation":

```{r}
model_std <- lm(
  scale(demlevel) ~ scale(trust) + bloc + scale(age) + scale(educ),
  data = swe
)

round(coef(model_std), 3)
```

`scale()` subtracts the mean and divides by the standard deviation. Now every continuous coefficient is in the same units and the sizes can be compared directly — trust does about twenty times the work of education. The cost is that the coefficients no longer mean anything in the units of the data, which is why this is a supplement rather than a replacement.

## Predicted values

The most useful thing you can show a reader is what the model expects to happen.

```{r}
predictions <- predict_response(model3, terms = "trust [-2:2 by=0.5]")

predictions
```

`predict_response()` from `ggeffects` computes predicted outcomes across the range of one variable, holding the others at representative values — it tells you which, underneath the table. `terms` names the variable to vary, and the bracket notation `[-2:2 by=0.5]` specifies the values to predict at.

```{r}
#| label: fig-pred
#| fig-height: 3.4
plot(predictions) +
  labs(
    x = "Trust in institutions (standard deviations from the mean)",
    y = "Predicted rating of Swedish democracy",
    title = NULL
  ) +
  theme_bw()
```

The shaded band is the 95% confidence interval around the prediction.

This says the same thing as the coefficient `r fmt(coef(model3)[2])`, but in a form a reader can use: someone two standard deviations below average on trust is predicted to rate Swedish democracy around `r fmt(predictions$predicted[predictions$x == -2])`, and someone two above at around `r fmt(predictions$predicted[predictions$x == 2])`.

### A warning the model will not give you

Look at the top of that prediction table.

```{r}
predictions |>
  as.data.frame() |>
  filter(x >= 1.5) |>
  select(x, predicted, conf.low, conf.high) |>
  round(2)
```

The predicted value at the top of the range is `r fmt(predictions$predicted[predictions$x == 2])`, and the upper end of its confidence interval is `r fmt(predictions$conf.high[predictions$x == 2])`.

**The scale stops at 10.** The model has produced a prediction that cannot happen, and it did so without complaint.

This is session 4 arriving in the middle of a regression. OLS fits a straight line, and a straight line does not know that the outcome is bounded. Extend it far enough in either direction and it will predict impossible values — a rating of 12, a negative income, a probability of 1.4.

Three things follow.

- **Do not predict outside the range of your data.** The model has no information there.
- **Check predictions against the limits of the scale**, as a matter of routine.
- **When a bounded outcome is the whole problem**, a straight line is the wrong model. That is what session 15 is about.

## Formatting a table or figure

The rules from session 5 apply, plus a few specific to model output.

- **Name variables in words.** "Trust in institutions", not `trust`.
- **Say what one unit is**, in the caption or a note. A coefficient without its units cannot be read.
- **Round sensibly.** Two decimal places is almost always enough. `1.0381823` claims a precision that the standard error of 0.04 flatly denies.
- **Report *n* for every model**, because listwise deletion makes it differ between columns. Ours run from `r nobs(model1)` down to `r nobs(model3)`.
- **Keep the models in a sensible order** — usually simplest first — and keep the coefficient order the same across columns.
- **Do not report everything.** A table with forty control variables hides the three that matter. Put controls in an appendix and say so.
- **Text on a figure should be about the size of the text around it.**

::: {.reporting}
**How to write it up.**

> Table 1 reports OLS models of how democratic respondents consider Sweden to be, on a 0-10 scale. Trust in institutions is the dominant predictor: a one standard deviation increase corresponds to a `r fmt(coef(model3)[2])`-point higher rating (95% CI [`r fmt(confint(model3)[2, 1])`, `r fmt(confint(model3)[2, 2])`]). Having voted for the right-wing bloc is associated with a `r fmt(abs(coef(model3)[3]))`-point lower rating, an estimate that is only marginally distinguishable from zero once trust is accounted for (*p* = `r fmt(coef(summary(model3))[3, 4], 3)`). Education has a small positive association and age none. The full model accounts for `r fmt(100 * summary(model3)$r.squared, 1)`% of the variance (*n* = `r nobs(model3)`).

Walk the reader through the table rather than repeating it. Mistakes to avoid:

- **Reproducing the table in prose.** Say what it shows; do not read it aloud.
- **Reporting a coefficient without its units or its uncertainty.**
- **Describing a coefficient whose interval crosses zero as an effect.**
- **Comparing raw coefficients on different scales** as though the sizes were comparable.
- **Predicting outside the range of the data**, or reporting a prediction the scale cannot hold.
- **Giving four decimal places** to a quantity with a standard error in the second.
- **Reporting a different *n* in the text than in the table**, which is what happens when the table is regenerated and the prose is not.
:::

::: {.swirl}
**In the seminar.** `swirl` lesson **12 Regression Presentation**. Building a table with several models, renaming and ordering coefficients, coefficient plots, and predicted values. Around 30 minutes.
:::
